---
title: "RNA-seq differential expression report"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 150)
suppressPackageStartupMessages({
  library(DESeq2)
  library(ggplot2)
  library(ggrepel)
  library(pheatmap)
  library(dplyr)
  library(DT)
})

dds      <- readRDS(snakemake@input[["dds"]])
vsd      <- readRDS(snakemake@input[["vst"]])
res      <- read.delim(snakemake@input[["results"]], stringsAsFactors = FALSE)
gsea     <- read.delim(snakemake@input[["gsea"]], stringsAsFactors = FALSE)
contrast <- unlist(snakemake@params[["contrast"]])
alpha    <- as.numeric(snakemake@params[["alpha"]])

label <- paste(contrast[2], "vs", contrast[3])
n_sig <- sum(res$padj < alpha, na.rm = TRUE)
```

## Summary

Contrast: **`r label`** (factor `r contrast[1]`), tested at padj < `r alpha`.

- Samples: **`r ncol(dds)`** (`r paste(table(colData(dds)[[contrast[1]]]), names(table(colData(dds)[[contrast[1]]])), collapse = ", ")`)
- Genes retained after filtering: **`r nrow(dds)`**
- Genes at padj < `r alpha`: **`r n_sig`**
- Median library size: **`r format(round(median(colSums(counts(dds)))), big.mark = ",")`** assigned counts

```{r caveat, results='asis'}
if (n_sig < 50) {
  cat("\n> With fewer than 50 significant genes, treat individual hits as",
      "hypotheses rather than findings. Check the per-sample counts for any",
      "gene before building on it — a large fold change driven by near-zero",
      "counts in one group is common at small sample sizes.\n")
}
```

## Sample-level diagnostics

### Principal components

```{r pca}
group <- contrast[1]
pca <- plotPCA(vsd, intgroup = group, returnData = TRUE)
pct <- round(100 * attr(pca, "percentVar"))
ggplot(pca, aes(PC1, PC2, colour = .data[[group]], label = name)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, show.legend = FALSE) +
  labs(x = paste0("PC1: ", pct[1], "% variance"),
       y = paste0("PC2: ", pct[2], "% variance")) +
  theme_bw()
```

Separation by condition here means treatment is the dominant source of variance.
Absence of separation does not by itself invalidate a short list of consistently
changed genes — it means condition is not the largest axis of variation, which
is common when between-animal or dissection variability is high.

### Sample-to-sample distance

```{r distheat, fig.height=5}
d <- as.matrix(dist(t(assay(vsd))))
ann <- as.data.frame(colData(dds)[, group, drop = FALSE])
pheatmap(d, annotation_col = ann, annotation_row = ann,
         main = "Euclidean distance on VST counts", silent = FALSE)
```

### Library size and dispersion

```{r libsize, fig.height=3.5}
ls <- data.frame(sample = colnames(dds),
                 counts = colSums(counts(dds)),
                 group = colData(dds)[[group]])
ggplot(ls, aes(reorder(sample, counts), counts, fill = group)) +
  geom_col() + coord_flip() +
  labs(x = NULL, y = "Assigned counts") + theme_bw()
```

```{r disp, fig.height=4.5}
plotDispEsts(dds)
```

The dispersion plot should show gene-wise estimates scattered around a smooth
decreasing trend, with shrunken values pulled toward it. A poor fit here means
the downstream p-values are unreliable.

## Differential expression

### Volcano

```{r volcano, fig.height=6}
v <- res %>%
  filter(!is.na(padj)) %>%
  mutate(
    direction = case_when(
      padj < alpha & log2FoldChange > 0 ~ "up",
      padj < alpha & log2FoldChange < 0 ~ "down",
      TRUE ~ "ns"
    ),
    lab = ifelse(rank(padj) <= 20 & padj < alpha,
                 ifelse(gene_symbol == "", gene_id, gene_symbol), NA)
  )

ggplot(v, aes(log2FoldChange, -log10(pvalue), colour = direction)) +
  geom_point(alpha = 0.5, size = 1.2) +
  geom_text_repel(aes(label = lab), size = 3, max.overlaps = 30,
                  show.legend = FALSE) +
  scale_colour_manual(values = c(up = "#c0392b", down = "#2c7fb8", ns = "grey75")) +
  labs(title = label, x = "log2 fold change (shrunken)") +
  theme_bw()
```

### Top genes

```{r toptable}
top <- res %>%
  filter(!is.na(padj)) %>%
  arrange(padj) %>%
  head(50) %>%
  mutate(across(c(baseMean, log2FoldChange, lfcSE), ~ round(.x, 3)),
         across(c(pvalue, padj), ~ signif(.x, 3)))
datatable(top, options = list(pageLength = 15, scrollX = TRUE), rownames = FALSE)
```

### Expression of top hits

```{r hitheat, fig.height=7}
sig <- res %>% filter(!is.na(padj)) %>% arrange(padj) %>% head(30)
if (nrow(sig) >= 2) {
  m <- assay(vsd)[sig$gene_id, , drop = FALSE]
  rownames(m) <- ifelse(sig$gene_symbol == "", sig$gene_id, sig$gene_symbol)
  pheatmap(m, scale = "row", annotation_col = ann,
           main = "Top genes by padj (VST, row-scaled)",
           fontsize_row = 8)
} else {
  cat("Fewer than two genes passed filtering; heatmap skipped.")
}
```

Row-scaled, so this shows relative pattern rather than absolute level. Genes
whose apparent effect comes from a single sample are visible immediately here.

## Gene set enrichment

Rank-based GSEA over the full gene list, ranked by
`sign(log2FC) * -log10(p)`. This uses every gene rather than only those passing
the significance threshold, which matters when that list is short.

```{r gseaplot, fig.height=9}
knitr::include_graphics(file.path(getwd(), snakemake@input[["gsea_plot"]]))
```

```{r gseatable}
gt <- gsea %>%
  filter(!is.na(padj)) %>%
  arrange(padj) %>%
  select(collection, pathway, NES, pval, padj, size) %>%
  head(50) %>%
  mutate(NES = round(NES, 3), across(c(pval, padj), ~ signif(.x, 3)))
datatable(gt, options = list(pageLength = 15, scrollX = TRUE), rownames = FALSE)
```

## Provenance

```{r prov}
cat("Design:      ", deparse(design(dds)), "\n")
cat("Contrast:    ", paste(contrast, collapse = " / "), "\n")
cat("Generated:   ", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n\n")
sessionInfo()
```
